---
title: "January 2023 Viral & Vaccine News"
author: "Pragati Prasad"
date: "2023-02-07"
slug: news-jan23
categories: ["News"]
output: html_document
---

This month in virus and vaccine news... 

 - China's lifting of COVID-19 restrictions is being monitored abroad.
 - Several other countries are following suit in loosening COVID-19 mitigations: Hong Kong, Japan, Spain, Italy, and South Korea.
 - Israel and EU conduct independent reviews of Pfizer vaccine and don't find a link to strokes.
 - Uganda announces the end of recent Ebola epidemic.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyRSS)
library(tidyverse)
library(lubridate)
library(stringr)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
file <- list.files(pattern = "2023-01")
start_date = as.Date("2023-01-01")
end_date = as.Date("2023-01-31")

terms <- "outbreak|virus|vaccin"

load(file)

get_news <- function(rss){
  r <- tidyfeed(
    feed = rss,
    config = list(),
    clean_tags = TRUE,
    list = FALSE,
    parse_dates = TRUE
  )
  
  final <- r[r %>%
    mutate_all(funs(str_detect(., terms)))  %>% 
    reduce(`|`) ,] 
    # select(where(~n_distinct(.) > 1))
  
  if("entry_last_updated" %in% colnames(final)){
    final$item_pub_date = final$entry_last_updated
  } else{
    final$item_pub_date <- if_else(is.na(final$item_pub_date), final$feed_pub_date, final$item_pub_date)
  }
  
  final <- final %>%
    mutate(item_pub_date = str_extract(item_pub_date, "^.{10}")) %>%
    mutate(item_pub_date = as.Date(item_pub_date)) 

  return(final)
}

```

### United Nations


```{r, echo=FALSE, warning=FALSE, message=FALSE}
un = rbind(if(exists("un")) un , 
           get_news(rss = "https://news.un.org/feed/subscribe/en/news/topic/health/feed/rss.xml")) %>%
  filter(duplicated(item_link) == FALSE) %>%
  filter(item_pub_date >= start_date, item_pub_date <= end_date) %>%
  mutate(item_pub_date = format(item_pub_date, "%B %d")) %>% 
  arrange(item_pub_date)

un_list <- c()
  
for (i in 1:nrow(un)){
  un_list <- c(un_list, 
               paste0("[", un$item_pub_date[i], ": ", un$item_title[i], "]", "(",un$item_link[i],")"))
}
  
newstring<-paste(un_list, "", collapse = "", sep = "  \n\n")
```

`r newstring`

### CDC

```{r, echo=FALSE, warning=FALSE, message=FALSE}
cdc <- rbind(#if(exists("cdc")) cdc,
             get_news(rss = "https://tools.cdc.gov/podcasts/feed.asp?feedid=183"))%>%
  filter(duplicated(entry_url) == FALSE) %>%
  filter(item_pub_date >= start_date, item_pub_date <= end_date) %>%
  mutate(item_pub_date = format(item_pub_date, "%B %d")) %>% 
  arrange(item_pub_date)

cdc_list <- c()

for (i in 1:nrow(cdc)){
  cdc_list <- c(cdc_list, paste0("[", cdc$item_pub_date[i], ": ", cdc$entry_title[i], "]", "(",cdc$entry_url[i],")"))
}

newstring<-paste(cdc_list, "", collapse = "", sep = "  \n\n")
```

`r newstring`

### New York Times

```{r, echo=FALSE, warning=FALSE, message=FALSE}
nyt <- rbind(if(exists("nyt")) nyt,
             get_news(rss = "https://rss.nytimes.com/services/xml/rss/nyt/Health.xml"))%>%
  filter(duplicated(item_link) == FALSE) %>%
  filter(item_pub_date >= start_date, item_pub_date <= end_date) %>%
  mutate(item_pub_date = format(item_pub_date, "%B %d")) %>% 
  arrange(item_pub_date)

nyt_list <- c()

for (i in 1:nrow(nyt)){
  nyt_list <- c(nyt_list, paste0("[", nyt$item_pub_date[i], ": ", nyt$item_title[i], "]", "(",nyt$item_link[i],")"))
}

newstring<-paste(nyt_list, "", collapse = "", sep = "  \n\n")

```

`r newstring`

### CNN

```{r, echo=FALSE, warning=FALSE, message=FALSE}
cnn <- rbind(if(exists("cnn")) cnn,
             get_news(rss = "http://rss.cnn.com/rss/cnn_health.rss")) %>%
  filter(duplicated(item_link) == FALSE)  %>%
  mutate(item_title = ifelse(is.na(item_title), item_description, item_title)) %>%
  filter(!is.na(item_title) & !is.na(item_link) & !is.na(item_description) & !is.na(item_guid)) %>%
  filter(item_pub_date >= start_date & item_pub_date <= end_date) %>%
  mutate(item_pub_date = format(item_pub_date, "%B %d")) %>% 
  arrange(item_pub_date) 
  
cnn_list <- c()

for (i in 1:nrow(cnn)){
  cnn_list <- c(cnn_list, paste0("[", cnn$item_pub_date[i], ": ", cnn$item_title[i], "]", "(",cnn$item_link[i],")"))
}

newstring<-paste(cnn_list, "", collapse = "", sep = "  \n\n")

if(nrow(cnn) == 0){
  newstring = "None."
}
```

`r newstring`


### STAT News

```{r, echo=FALSE, warning=FALSE, message=FALSE}
stat <- rbind(if(exists("stat")) stat,
              get_news(rss = "https://www.statnews.com/feed/")) %>%
  filter(duplicated(item_link) == FALSE) %>%
  filter(item_pub_date >= start_date, item_pub_date <= end_date) %>%
  mutate(item_pub_date = format(item_pub_date, "%B %d")) %>% 
  arrange(item_pub_date) 

stat_list <- c()

for (i in 1:nrow(stat)){
  stat_list <- c(stat_list, paste0("[", stat$item_pub_date[i], ": ", stat$item_title[i], "]", "(",stat$item_link[i],")"))
}

newstring<-paste(stat_list, "", collapse = "", sep = "  \n\n")

```

`r newstring`

### FOX News 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
fox <- rbind(if(exists("fox")) fox,
             get_news(rss = "https://moxie.foxnews.com/google-publisher/health.xml")) %>%
  filter(duplicated(item_guid) == FALSE) %>%
  filter(item_pub_date >= start_date, item_pub_date <= end_date) %>%
  mutate(item_pub_date = format(item_pub_date, "%B %d")) %>% 
  arrange(item_pub_date) 

fox_list <- c()

for (i in 1:nrow(fox)){
  fox_list <- c(fox_list, paste0("[", fox$item_pub_date[i], ": ", fox$item_title[i], "]", "(",fox$item_guid[i],")"))
}

newstring<-paste(fox_list, "", collapse = "", sep = "  \n\n")
```

`r newstring`


### Reuters

```{r, echo=FALSE, warning=FALSE, message=FALSE}
reu <- rbind(if(exists("reu")) reu,
             get_news(rss = "https://www.reutersagency.com/feed/?best-topics=health&post_type=best")) %>%
  filter(duplicated(item_link) == FALSE) %>%
  filter(item_pub_date >= start_date, item_pub_date <= end_date) %>%
  mutate(item_pub_date = format(item_pub_date, "%B %d")) %>% 
  arrange(item_pub_date) 

reu_list <- c()

for (i in 1:nrow(reu)){
  reu_list <- c(reu_list, paste0("[", reu$item_pub_date[i], ": ", reu$item_title[i], "]", "(",reu$item_link[i],")"))
}

newstring<-paste(reu_list, "", collapse = "", sep = "  \n\n")

if(nrow(reu) == 0){
  newstring = "None."
}
```

`r newstring`


### NIH

```{r, echo=FALSE, warning=FALSE, message=FALSE}
nih <- rbind(if(exists("nih")) nih,
             get_news(rss = "https://www.techtransfer.nih.gov/taxonomy/term/232/feed")) %>%
  filter(duplicated(item_link) == FALSE) %>%
  filter(item_pub_date >= start_date, item_pub_date <= end_date) %>%
  mutate(item_pub_date = format(item_pub_date, "%B %d")) %>% 
  arrange(item_pub_date) 

nih_list <- c()

for (i in 1:nrow(nih)){
  nih_list <- c(nih_list, paste0("[", nih$item_pub_date[i], ": ", nih$item_title[i], "]", "(",nih$item_link[i],")"))
}

newstring<-paste(nih_list, "", collapse = "", sep = "  \n\n")
```

`r newstring`


### Al Jazeera

```{r, echo=FALSE, warning=FALSE, message=FALSE}
alj <- rbind(if(exists("alj")) alj,
             get_news(rss = "https://www.aljazeera.com/xml/rss/all.xml")) %>%
  filter(duplicated(item_link) == FALSE) %>%
  filter(item_pub_date >= start_date, item_pub_date <= end_date) %>%
  mutate(item_pub_date = format(item_pub_date, "%B %d")) %>% 
  arrange(item_pub_date) 

alj_list <- c()

for (i in 1:nrow(alj)){
  alj_list <- c(alj_list, paste0("[", alj$item_pub_date[i], ": ", alj$item_title[i], "]", "(",alj$item_link[i],")"))
}

newstring<-paste(alj_list, "", collapse = "", sep = "  \n\n")
```

`r newstring`

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(lubridate)
save(alj, cdc, cnn, fox, nih, nyt, reu, stat, un, file = paste0(today(), "_news.RData"))
```