---
title: CHIKV Model
author: Pragati Prasad
date: '2019-10-28'
slug: chikv-model
categories: []
tags: []
---

Based on this:

Yakob L, Clements AC. A mathematical model of chikungunya dynamics and control: the major epidemic on Réunion Island. PLoS One. 2013;8(3):e57448. doi:10.1371/journal.pone.0057448

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3590184/


```{r}
suppressMessages(library(EpiModel))

```

```{r}

chikv <-function(t, t0, parms) {
  with(as.list(c(t0, parms)), {
    
    num <- s.num + e.num + i.num + ia.num + r.num 
    num.m <- x.num + y.num + z.num

    dS <- -1*beta1*(s.num)*(z.num / num.m)
    dE <- (beta1*(s.num)*(z.num / num.m)) - (lambda1*(e.num))
    dI <- (theta*lambda1*(e.num)) - (gamma*(i.num))
    dIa <- ((1 - theta)*lambda1*(e.num)) - (gamma*(ia.num))
    dR <- gamma*((i.num + ia.num))
    
    dX <- (mu*x.num) - (beta2*(x.num)*((i.num+ia.num) / num)) - (mu*(x.num))
    dY <- (beta2*(x.num)*((i.num+ia.num) / num)) - (lambda2*(y.num)) - (mu*(y.num))
    dZ <- (lambda2*(y.num)) - (mu*(z.num))

    list(c(dS,
           dE,
           dI,
           dIa,
           dR,
           dX,
           dY,
           dZ,
           se.flow <- (beta1*s.num*(z.num / num.m)),
           ei.flow <- (lambda1*e.num),
           ir.flow<- gamma*(i.num + ia.num),
           xy.flow <- (beta2*x.num*((i.num + ia.num) / num)),
           yz.flow <- (lambda2*y.num)
    ))
  })
}

```

Population = 787,836 - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2528011/

Mosquito population = 9 million? - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6069545/
Brazil has 10-250 million mosquitos, but I use 9 million for Reunion Island

```{r}
param <-param.dcm(beta1 = 0.14,
                  beta2 = 0.4,
                  theta = 0.97,
                  lambda1 = 0.5,
                  lambda2 = 0.5,
                  gamma = 0.25,
                  mu = 0.05)
                  #omega = 0.25)
init <-init.dcm(s.num = 787836,
                e.num = 0,
                i.num = 0,
                ia.num = 0,
                r.num = 0,
                x.num = 3300000,
                y.num = 3300000,
                z.num = 3300000,
                se.flow = 0,
                ei.flow = 0,
                ir.flow = 0,
                xy.flow = 0,
                yz.flow = 0)
control <-control.dcm(nsteps =365, new.mod = chikv) 

sim <- dcm(param, init, control)

sim <- mutate_epi(sim, num = s.num + e.num + i.num + ia.num + r.num)
sim <- mutate_epi(sim, 
                  symp.prev = i.num / num,
                  se.prop = se.flow / num,
                  ei.prop = ei.flow / num,
                  ir.prop = ir.flow / num
)

plot(sim,y=c("s.num", "i.num", "e.num", "r.num"),legend="full")
plot(sim,y=c("i.num"),legend="full")
plot(sim,y=c("symp.prev"),legend="full")

plot(sim,y=c("se.flow", "ei.flow", "ir.flow"),legend="full")
plot(sim,y=c("se.prop", "ei.prop", "ir.prop"),legend="full",ylim=c(0, 0.08))
```

```{r, echo=FALSE}
df<-as.data.frame(sim)
print("Peak Incidence")
max(df$se.prop) 

print("Peak Prevalence")
max(df$symp.prev) 
```
Peak Incidence here is not consistent with the paper's findings (5.3%).
But we've only done a deterministic analysis here with static parameters.

```{r}
plot(sim,y=c("x.num", "y.num", "z.num"),legend="full")
plot(sim,y=c("xy.flow", "yz.flow"),legend="full")
```


However, it doesn't make sense to me that only uninfected mosquitos (in the X compartment) 
would be breeding. Wouldn't mosquitos in all compartments procreate?

I change this below

```{r}

chikv <-function(t, t0, parms) {
  with(as.list(c(t0, parms)), {
    
    num <- s.num + e.num + i.num + ia.num + r.num 
    num.m <- x.num + y.num + z.num

    dS <- -1*beta1*(s.num)*(z.num / num.m)
    dE <- (beta1*(s.num)*(z.num / num.m)) - (lambda1*(e.num))
    dI <- (theta*lambda1*(e.num)) - (gamma*(i.num))
    dIa <- ((1 - theta)*lambda1*(e.num)) - (gamma*(ia.num))
    dR <- gamma*((i.num + ia.num))
    
    dX <- (mu*num.m) - (beta2*(x.num)*((i.num+ia.num) / num)) - (mu*(x.num))
    dY <- (beta2*(x.num)*((i.num+ia.num) / num)) - (lambda2*(y.num)) - (mu*(y.num))
    dZ <- (lambda2*(y.num)) - (mu*(z.num))

    list(c(dS,
           dE,
           dI,
           dIa,
           dR,
           dX,
           dY,
           dZ,
           se.flow <- (beta1*s.num*(z.num / num.m)),
           ei.flow <- (lambda1*e.num),
           ir.flow<- gamma*(i.num + ia.num),
           xy.flow <- (beta2*x.num*((i.num + ia.num) / num)),
           yz.flow <- (lambda2*y.num)
    ))
  })
}

```

```{r}
param <-param.dcm(beta1 = 0.14,
                  beta2 = 0.4,
                  theta = 0.97,
                  lambda1 = 0.5,
                  lambda2 = 0.5,
                  gamma = 0.25,
                  mu = 0.05)
                  #omega = 0.25)
init <-init.dcm(s.num = 787836,
                e.num = 0,
                i.num = 0,
                ia.num = 0,
                r.num = 0,
                x.num = 3300000,
                y.num = 3300000,
                z.num = 3300000,
                se.flow = 0,
                ei.flow = 0,
                ir.flow = 0,
                xy.flow = 0,
                yz.flow = 0)
control <-control.dcm(nsteps =365, new.mod = chikv) 

sim <- dcm(param, init, control)

sim <- mutate_epi(sim, num = s.num + e.num + i.num + ia.num + r.num)
sim <- mutate_epi(sim, 
                  symp.prev = i.num / num,
                  se.prop = se.flow / num,
                  ei.prop = ei.flow / num,
                  ir.prop = ir.flow / num
)

plot(sim,y=c("s.num", "i.num", "e.num", "r.num"),legend="full")
plot(sim,y=c("i.num"),legend="full")
plot(sim,y=c("symp.prev"),legend="full")

plot(sim,y=c("se.flow", "ei.flow", "ir.flow"),legend="full")
plot(sim,y=c("se.prop", "ei.prop", "ir.prop"),legend="full",ylim=c(0, 0.08))
```

```{r, echo=FALSE}
df<-as.data.frame(sim)
print("Peak Incidence")
max(df$se.prop) 

print("Peak Prevalence")
max(df$symp.prev) 
```

Peak prevalence and peak incidence are reduced here - because we introduce more 
uninfected susceptibles than in the prior model if I assume that all mosquitos 
breed to create suseceptible mosquitos.
